Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-27 manuscript/manuscript.tex
Manuscript update
Removing the computed fields reference and analysis.

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.7\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

@@

\jy{The plot sizes seem to be fixed in the generation code. That's why
  the fonts are small when scaled. I'd generate them with approximately
  the expected sizes in the manuscript. I'd also use aspect ratio
  0.618 (golden ratio).
}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.7\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.7\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint types and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

@@

\jy{Can we lighten the color of the bar to make the pertentages more
  legible?}

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes (two fields: \texttt{zip\_codes} and \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
37,946 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors; 
not insignificant. If you sort the invalid entries by Agency, 
the distribution by percentage mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\begin{table}[tbp]
    \centering
    \caption{Comparison of Top Ten Zip Codes Lists}
	    \begin{tabular}{@{}lS[table-format=6.0,round-mode=places,
	    round-precision=0]c@{\hskip 0.5cm}@{}lS[table-format=6.0,
	    round-mode=places,round-precision=0]c@{}}
		\toprule
	 	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	      \cmidrule(r){1-3} \cmidrule(l){4-6}
	      \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} 
	      & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
	      \midrule
	        11275 & 104556 & FALSE & 10466 & 104562 & TRUE \\
	        12420 & 27503 & TRUE & 10023 & 27972 & TRUE \\
	        12428 & 26564 & TRUE & 10031 & 25548 & TRUE \\
	        10935 & 25508 & FALSE & 10457 & 25066 & TRUE \\
	        10934 & 23448 & FALSE & 10453 & 24752 & TRUE \\
	        10931 & 22381 & TRUE & 10456 & 24751 & TRUE \\
	        10930 & 22121 & TRUE & 10452 & 22527 & TRUE \\
	        17613 & 21963 & FALSE & 10025 & 21705 & TRUE \\
	        10936 & 21707 & FALSE & 10458 & 21689 & TRUE \\
	        11606 & 21435 & FALSE & 10032 & 20622 & TRUE \\
	      \bottomrule
	    	\end{tabular}
 	\label{tab:zipcodes}
\end{table}

Consider a case study with the NYC Office of Nightlife (ONL). Suppose 
that ONL wants to know, "What are the top 10 zip codes for Noise 
Complaints over the last two years?" ONL seeks to assess the impact 
of recent efforts promoting a vibrant nightlife in NYC while also 
easing tensions between bar and club owners. Assume this analysis uses 
the \texttt{zip\_codes} field, one of the computed fields known to have 
validity issues. The analysis is then repeated using the more accurate 
\texttt{incident\_zip} field. Finally, both analyses are validated 
against the USPS zip code database. The top 10 zip codes and their 
validation results are summarized in Table~\ref{tab:zipcodes}. Six out 
of the ten \texttt{zip\_codes} entries are invalid, closely matching 
the overall dataset's 58\% invalid rate. In contrast, the 
\texttt{incident\_zip} field is fully valid, consistent with the 
overall \texttt{incident\_zip} validation rate of 99\%.
It is possible that the computed \texttt{zip\_codes} may mean the zip
code of something differnent from the incident. Without proper
documentation, it is confusing to have it in the released data.

\paragraph{Police Precincts}
\label{sec:police-precincts}
A curious case arises when examining two nearly identical fields: 
\texttt{police\_precincts} and \texttt{police\_precinct}. Both fields 
are among the computed fields previously mentioned. Using the 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings}, it is possible to determine the valid police 
precincts. Upon validation, both \texttt{police\_precincts} and 
\texttt{police\_precinct} contain about 35\% invalid entries. However, 
the invalid entries are not identical. The \texttt{police\_precincts} 
field has 2,171,864 invalid entries (35\%), while the 
\texttt{police\_precinct} field has 2,171,778 invalid entries (also 
35\%), but they do not match.

\paragraph{Community Boards}
\label{sec:communityboards}
Community Boards are the most local, grassroots form of City government, 
serving as a vital connection between communities, elected officials, 
and City agencies. They often act as a measure of the equity of City 
services across the five boroughs. In the 2022-2023 dataset, there are 
27,276 invalid \texttt{community\_board} entries, representing 0.43\% 
of non-blank data. The distribution of these invalid entries by Agency 
is not consistent with the overall SR Agency 
distribution. This suggests that issues may exist within specific key 
Agencies, such as the Taxi \& Limousine Commission (TLC) and the 
Department of Parks and Recreation (DPR).

\paragraph{Community Districts}
\label{sec:communitydistrict}
Another of the computed fields is \texttt{community\_districts}. 
Community Districts define the geographic boundaries for the Community 
Boards but differ in that they are used by the Department of City 
Planning (DCP) for environmental, socio-economic, and demographic 
purposes. They represent a geographical division rather than a City 
government division. Due to the format of the \texttt{community\_district} 
data, it is not possible to directly establish validity. However, the 
dataset contains 72 unique entries, while only 59 valid Community 
Districts exist.

@@

\begin{figure}[tbp]
	 \centering
 	 \includegraphics[width = 0.7\textwidth]{negative_duration_SR_violin.pdf}
 \caption{Distribution of Negative Durations}
 \label{fig:negative-duration-violin}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width = 0.7\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed resolution action update date(s) over 30 days.}
  \label{fig:resolution-violin}
\end{figure}

\jy{Are these from 2-year or 10-year period? Need to clarify in
  Section 2.}

@@

\textbf{borough \& borough\_boundaries:} These fields are 98.3\% matches. 
This ``near-duplicate'' relationship creates a challenge in determining 
which field is correct when discrepancies arise.

@@

\textbf{incident\_zip \& zip\_codes:} The \texttt{zip\_codes} field, 
one of the computed fields, was found to have 56\% invalid entries, 
while the \texttt{incident\_zip} field has a 99.93\% accuracy rate. 
We recommend eliminating or masking the \texttt{zip\_codes} field 
until its accuracy can be improved.

\textbf{police\_precinct \& police\_precincts:} These two computed 
fields are 99.9\% matches. Since the computational process is unknown, 
it is unclear which field is more accurate. However, having two fields 
with near-duplicity suggests that only one is necessary.

@@

By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 12.3\%, which for this large dataset equates to 
a reduction of 395 Mb. A smaller file size means faster downloads, 
less storage impact, and simpler data analysis efforts.
Below is a list of duplicate and near-duplicate fields. There are 
challenges with these proposed deletions, especially with the 
``near-duplicate'' fields. For example, while the \texttt{park\_borough} 
is a complete duplicate of the \texttt{borough} field, the 
\texttt{borough\_boundary} field is only 98.3\% similar. Although losing 
1.7\% of the data is small, it still amounts to 110,715 non-matching 
occurrences. The authors propose deleting the following fields:

@@

    \item \textbf{police\_precinct}: This field has a 99.95\% match with 
    \texttt{police\_precincts}. Removing one would result in minimal data loss.

    \item \textbf{borough\_boundaries (computed field)}: A 98.3\% match 
    with the \texttt{borough} field.

    \item \textbf{cross\_street\_1 \& 2 and intersection\_street\_1 \& 2}: 
    An 88\% match between these pairs. We recommend deleting the 
    \texttt{intersection\_street\_1 \& 2} fields while acknowledging some 
    data loss.

  \item \textbf{zip\_codes (computed field)}:
    This field has a 58\% error rate and can lead to significant 
    analytical errors. We recommend deleting this field.
\end{itemize}

@@

  \item As mentioned earlier, six \texttt{@computed} fields that are
    present in the data but not included in the Data Dictionary need
    to be documented.
\end{itemize}

@@

\subsection{Principles}
\paragraph{Principle 1: Ensuring Consistency Across Agencies and Time}
A key challenge in government-released datasets is the 
harmonization of data fields across various agencies and 
maintaining this consistency over time. As seen in the NYC 
311 data, discrepancies in field names, formats, and definitions, 
such as those between \texttt{police\_precinct} and 
\texttt{police\_precincts}, can introduce significant 
inconsistencies. To address this, agencies should adopt 
standardized naming conventions and formats across all datasets. 
Regular audits and data harmonization processes should be 
implemented to ensure consistency across long-term datasets, 
particularly as agency structures or reporting standards evolve. 
For instance, ensuring that historical datasets reflect the same 
field naming conventions over a span of years can prevent 
inconsistent analyses.

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes (e.g., 58\% invalid \texttt{zip\_codes} entries) and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields like 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data.

\paragraph{Principle 3: Optimizing Storage and Representation Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. By eliminating redundant fields and 
encoding categorical variables, agencies can significantly reduce 
file sizes and improve the efficiency of data analysis. In the NYC 
311 dataset, for example, redundant fields such as \texttt{borough} 
and \texttt{park\_borough} could be consolidated without any data 
loss, saving space and reducing file size by up to 12.3\%. Another 
example is the \texttt{taxi\_company\_borough} field, which is 99.94\% 
blank and should be evaluated to determine if its inclusion is 
necessary. Similarly, categorical variables like \texttt{complaint\_type} 
could be encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves storage space 
but also enhances the performance of queries and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government-released datasets. The NYC 311 data revealed several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields. 
Updating the Data Dictionary to include descriptions of all fields, 
especially computed fields like \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
be made to ensure that any changes in data structure or content are 
reflected. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight or noon likely stem from automated processes that 
distort the accurate capture of SR durations. By implementing 
real-time validation tools, governments can prevent such anomalies 
from entering the dataset. Automated systems can be designed to 
flag inconsistencies, incorrect values, or illogical sequences, such 
as invalid \texttt{incident\_zip} values or illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user-friendly formats. For example, in the NYC 
311 dataset, fields like \texttt{@computed\_region\_xxxx\_xxxx} and 
the distinction between \texttt{incident\_zip} and \texttt{zip\_codes} 
can be confusing for non-expert users. These fields should be clearly 
documented, explaining how they are calculated and how they differ. 
Additionally, governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous closed dates) are discovered and explaining 
how such issues will be resolved. By ensuring transparency and 
keeping the public informed, governments foster trust and promote 
wider use of datasets for research, civic engagement, and innovation.

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reported in
this paper are available [FixMe].
\jy{need to think: we need to keep a copy of the downloaded data;
  if the data portal changes and our code may not work.}

>>>>

\begin{figure}[tbp]
	\centering
  	\includegraphics[width = \textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

@@

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

\begin{figure}[tbp]
 \centering
  \includegraphics[width = \textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint types and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

@@

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors; 
not insignificant. If you sort the invalid entries by Agency, 
the distribution by percentage mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

@@

\begin{figure}[tbp]
	 \centering
 	 \includegraphics[width=\textwidth]{negative_duration_SR_violin.pdf}
 \caption{Distribution of Negative Durations}
 \label{fig:negative-duration-violin}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed resolution action update date(s) over 30 days.}
  \label{fig:resolution-violin}
\end{figure}

@@

@@

@@

By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 39.5\%, which for this large dataset equates to 
a reduction of 1.4 Gb, which is significant. A smaller file size means faster downloads, 
less storage impact, and simpler data analysis efforts.
Below is a list of duplicate and near-duplicate fields.

@@

    \item \textbf{cross\_street\_1 \& 2 and intersection\_street\_1 \& 2}: 
    An 88\% match between these pairs. We recommend deleting the 
    \texttt{intersection\_street\_1 \& 2} fields while acknowledging the
    concomitant data loss.

\end{itemize}

@@

  \item As mentioned earlier, six \texttt{@computed} fields that are
    present in the data but not included in the Data Dictionary need
    to be documented and perhaps labelled as ``experimental'' or
    ``not for official use''.
\end{itemize}

@@

\subsection{Principles}
\paragraph{Principle 1: Ensuring Consistency Across Agencies and Time}
A key challenge in government-released datasets is the 
harmonization of data fields across various agencies and 
maintaining this consistency over time. To address this, agencies should adopt 
standardized naming conventions and formats across all datasets. 
Regular audits and data harmonization processes should be 
implemented to ensure consistency across long-term datasets, 
particularly as agency structures or reporting standards evolve. 
For instance, ensuring that historical datasets reflect the same 
field naming conventions over a span of years can prevent 
inconsistent analyses.

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields like 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data.

\paragraph{Principle 3: Optimizing Storage and Representation Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. By eliminating redundant fields and 
encoding categorical variables, agencies can significantly reduce 
file sizes and improve the efficiency of data analysis. In the NYC 
311 dataset, for example, redundant fields such as \texttt{borough} 
and \texttt{park\_borough} could be consolidated without any data 
loss, saving space and reducing file size. Another 
example is the \texttt{taxi\_company\_borough} field, which is 99.94\% 
blank and should be evaluated to determine if its inclusion is 
necessary. Similarly, categorical variables like \texttt{complaint\_type} 
could be encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves storage space 
but also enhances the performance of queries and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government-released datasets. The NYC 311 data revealed several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields. 
As mentioned, updating the Data Dictionary to include descriptions of all fields, 
including the computed fields, e.g. \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
be made to ensure that any changes in data structure or content are 
reflected. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight or noon likely stem from automated processes that 
distort the accurate capture of SR durations. By implementing 
real-time validation tools, governments can prevent such anomalies 
from entering the dataset. Automated systems can be designed to 
flag inconsistencies, incorrect values, or illogical sequences, or 
illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user-friendly formats. Additionally, 
governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous closed dates) are discovered and explaining 
how such issues will be resolved. By ensuring transparency and 
keeping the public informed, governments foster trust and promote 
wider use of datasets for research, civic engagement, and innovation.

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reported in
this paper are available [FixMe].
